\textbf{ Normalized Gaussian Processes }

### 2020-11-29 tex/ms.tex
new fig
more figures

\ifdefined \includeAppendix

    \clearpage

    \appendix

    \section{Computing $\tilde{\pmb{\Sigma}}$}
    %
    Since the expectation is a linear operator, we can write Equation~(\ref{eq:SigmaTildeExp})
    as
    %
    \begin{align}
        \tilde{\pmb{\Sigma}}
         & =
        \frac{1}{\mu^2}
        \sum\limits_{n=0}^\infty
        (-1)^n (n + 1)
        \,
        \mathbb{E}\left[
            \epsilon^n
            \Bigg(
            \mu^2 \mathbf{j} \, \mathbf{j}^\top
            +
            \mu \, \mathbf{j}\mathbf{u}^\top \mathbf{L}^\top
            +
            \mu \, \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top
            +
            \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \, \mathbf{L}^\top
            \Bigg)
            \right]
        - \mathbf{j} \, \mathbf{j}^\top
        \nonumber \\[0.5em]
         & =
        \sum\limits_{n=0}^\infty
        \frac{(-1)^n}{\mu^{2 + n}}
        \,
        \left(
        \mu^2 \mathbf{P}_n
        +
        \mu \, \mathbf{Q}_n
        +
        \mu \, \mathbf{Q}_n^\top
        +
        \mathbf{R}_n
        \right)
        - \mathbf{j} \, \mathbf{j}^\top
        \quad,
    \end{align}
    %
    where we define the matrices
    %
    \begin{align}
        \mathbf{P}_n & \equiv \frac{(n + 1)}{K^n}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^n \mathbf{j} \, \mathbf{j}^\top \right]
        \nonumber                                                                                                                                                                                \\[0.5em]
        \mathbf{Q}_n & \equiv \frac{(n + 1)}{K^n}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^n \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top \right]
        \nonumber                                                                                                                                                                                \\[0.5em]
        \mathbf{R}_n & \equiv \frac{(n + 1)}{K^n}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^n \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \, \mathbf{L}^\top \right]
        \quad.
    \end{align}
    %
    In order to compute these matrices,
    we will make use of the expression for the
    $n^\mathrm{th}$ moment of the standard normal distribution:
    %
    \begin{align}
        \mathbb{E}\left[ u^n \right] & = g_n
    \end{align}
    %
    for scalar $u \sim \mathcal{N}(0, 1)$, where
    %
    \begin{align}
        g_n
         & =
        \begin{cases}
            \dfrac{n!}{2^\frac{n}{2} \left(\frac{n}{2}\right)!} & n \, \mathrm{even}
            \\
            0                                                   & n \, \mathrm{odd}
            \quad.
        \end{cases}
    \end{align}
    %
    \citep[e.g.,][]{Winkelbauer2012}.
    We can apply this relation to compute expectations
    of products of the components of a random standard normal vector $\mathbf{u}$:
    %
    \begin{align}
        \mathbb{E}\left[u_i \right]                  & = 0
        \\[1em]
        \mathbb{E}\left[u_i u_j\right]               & = g_2
        \\[1em]
        \mathbb{E}\left[u_i u_j u\right]             & = 0
        \\[1em]
        \mathbb{E}\left[u_i u_j u u_l\right]         & =
        g_4 \, \delta_{i, j} \delta_{k, l} \, \delta_{i, k}
        +
        g_2^2 \left(
        \delta_{i, j} \, \delta_{k, l} \, \bar{\delta}_{i, k}
        +
        \delta_{i, k} \, \delta_{j, l} \, \bar{\delta}_{i, j}
        +
        \delta_{i, l} \, \delta_{j, k} \, \bar{\delta}_{i, j}
        \right)
        \\[1em]
        \mathbb{E}\left[u_i u_j u u_l u_m\right]     & = 0
        \\[1em]
        \mathbb{E}\left[u_i u_j u u_l u_m u_n\right] & =
        ...
        \quad,
    \end{align}
    %
    where $\delta$ is the Kronecker delta and $\bar{\delta} \equiv 1 - \delta$.

    \subsection{The $\mathbf{P}$ expectation}

    Let us begin by computing the first several terms in $\mathbf{P}$.
        \mathbf{P}_0 & = \mathbb{E}\left[ \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \mathbf{j} \, \mathbf{j}^\top
    since $\mathbf{j} \, \mathbf{j}^\top$ is constant. Next,
        \mathbf{P}_1 & = \frac{2}{K}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right) \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{2}{K}\left(\mathbf{j}^\top \mathbf{L} \, \mathbb{E}\left[ \mathbf{u} \right]\right) \mathbf{j} \, \mathbf{j}^\top
                     & = \mathbf{0}
    since $\mathbb{E}\left[ \mathbf{u} \right] = \mathbf{0}$. The next term is
        \mathbf{P}_2 & = \frac{3}{K^2}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^2 \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{3}{K^2}\mathbb{E}\left[\left(\mathbf{j}^\top \mathbf{L} \,  \mathbf{u} \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \right)  \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{3}{K^2}\left(\mathbf{j}^\top \mathbf{L} \, \mathbb{E}\left[ \mathbf{u} \mathbf{u}^\top \right] \mathbf{L}^\top \, \mathbf{j} \right)  \mathbf{j} \, \mathbf{j}^\top
                     & = \frac{3}{K^2}\left(\mathbf{j}^\top \mathbf{L} \mathbf{L}^\top \, \mathbf{j} \right)  \mathbf{j} \, \mathbf{j}^\top
                     & = \frac{3}{K^2}\mathrm{sum}(\mathbf{\Sigma}) \mathbf{j} \, \mathbf{j}^\top
                     & = 3 \, m \, \mathbf{j} \, \mathbf{j}^\top
    where $m$ is the average of all entries in $\pmb{\Sigma}$ and
    we made use of the fact that $\mathbb{E}\left[ \mathbf{u} \mathbf{u}^\top\right] = \mathbf{I}$.
    Next,
        \mathbf{P}_3 & = \frac{4}{K^3}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^3 \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{4}{K^3}\mathbb{E}\left[\left(\mathbf{j}^\top \mathbf{L} \,  \mathbf{u} \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \mathbf{L} \,  \mathbf{u} \right) \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{4}{K^3}\left(\mathbf{j}^\top \mathbf{L} \,  \mathbb{E}\left[\mathbf{u} \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \mathbf{L} \,  \mathbf{u}\right] \right) \mathbf{j} \, \mathbf{j}^\top
                     & = \mathbf{0}
    where we made use of the fact that we may write the component at index $i$ of the vector within the expectation in the next-to-last line
    as
        \mathbb{E}\left[\mathbf{u} \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \mathbf{L} \,  \mathbf{u}\right]_{i} = \sum\limits_{k,l}(\mathbf{L}^\top \mathbf{j} \, \mathbf{j}^\top \mathbf{L})_{kl} u_i u u_l = 0
        \nonumber
    which is zero since it is the product of an odd number of random normal variables.
    Finally,
        \mathbf{P}_4 & = \frac{5}{K^4}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^4 \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{5}{K^4}\mathbb{E}\left[\left(\mathbf{j}^\top \mathbf{L} \,  \mathbf{u} \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \mathbf{L} \,  \mathbf{u} \, \mathbf{u}^\top  \mathbf{L}^\top \, \mathbf{j} \right) \mathbf{j} \, \mathbf{j}^\top \right]
                     & = \frac{5}{K^4}\left(\mathbf{j}^\top \mathbf{L} \, \mathbf{E} \, \mathbf{L}^\top \, \mathbf{j} \right) \mathbf{j} \, \mathbf{j}^\top
        \mathbf{E} \equiv \mathbb{E}\left[ \mathbf{u} \mathbf{u}^\top  \mathbf{A} \,  \mathbf{u} \mathbf{u}^\top \right]
    and
        \mathbf{A} \equiv \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \, \mathbf{L}
    The components of this matrix are
        E_{ij} & =
        \sum\limits_{k,l}A_{kl} \mathbb{E}(u_i u_j u u_l)
        \sum\limits_{k,l}A_{kl}
        \Big(
        g_4 \, \delta_{i, j} \, \delta_{k, l} \, \delta_{i, k}
        g_2^2 \big(
            \delta_{i, j} \, \delta_{k, l} \, \bar{\delta}_{i, k}
            \delta_{i, k} \, \delta_{j, l} \, \bar{\delta}_{i, j}
            \delta_{i, l} \, \delta_{j, k} \, \bar{\delta}_{i, j}
            \big)
        \Big)
        2 A_{ij} + \sum\limits A_{kk}
    where we used the fact that $\mathbf{A} = \mathbf{A}^\top$. We may thus write
        \mathbf{E} & =
        2 \, \mathbf{A} + \mathrm{tr}(\mathbf{A}) \mathbf{I}
        2 \, \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \, \mathbf{L} + \mathrm{sum}(\pmb{\Sigma}) \mathbf{I}
    Inserting this back into the expression for $\mathbf{P}_4$, we obtain
        \mathbf{P}_4 & =
        \frac{5}{K^4}
        \bigg(
        2 \, \big(\mathbf{j}^\top \mathbf{L} \, \mathbf{L}^\top \, \mathbf{j} \, \mathbf{j}^\top \, \mathbf{L} \, \mathbf{L}^\top \, \mathbf{j} \big)
        \mathrm{sum}(\pmb{\Sigma}) \big(\mathbf{j}^\top \mathbf{L} \, \mathbf{L}^\top \, \mathbf{j} \big)
        \bigg) \mathbf{j} \, \mathbf{j}^\top
        3 \, m^2 \mathbf{j} \, \mathbf{j}^\top
    The general term in this sequence is
        \mathbf{P}_n & =
            g_n (n + 1) \, m^\frac{n}{2} & n \, \mathrm{even}

    \subsection{The $\mathbf{Q}$ expectation}
    Let us again compute the first several terms of this expectation matrix.
        \mathbf{Q}_0 & = \mathbb{E}\left[ \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top \right]
                     & = \mathbf{L} \, \mathbb{E}\left[  \mathbf{u} \right] \, \mathbf{j}^\top
                     & = \mathbf{0}
        \mathbf{Q}_1 & = \frac{1}{K}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right) \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top \right]
                     & = \frac{1}{K}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right) \mathbf{L} \, \mathbf{u} \right] \, \mathbf{j}^\top
                     & = \frac{1}{K}\mathbb{E}\left[ \mathbf{L} \, \mathbf{u} (\mathbf{u}^\top \mathbf{L}^\top \mathbf{j}) \right] \, \mathbf{j}^\top
                     & = \frac{1}{K}\mathbb{E}\left[ \mathbf{L} \, \mathbf{u} \mathbf{u}^\top \mathbf{L}^\top  \right] \, \mathbf{j} \, \mathbf{j}^\top
                     & = \frac{1}{K}\mathbf{L} \, \mathbb{E}\left[  \mathbf{u} \mathbf{u}^\top   \right]  \, \mathbf{L}^\top \mathbf{j} \, \mathbf{j}^\top
                     & = \frac{1}{K}\mathbf{L} \, \mathbf{L}^\top \mathbf{j} \, \mathbf{j}^\top
                     & = \frac{1}{K}\pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top
    where we made liberal use of the fact that $\left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)$ is a scalar. Continuing,
        \mathbf{Q}_2 & = \frac{2}{K^2}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^2 \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top \right]
                     & =  \frac{2}{K^2}\mathbb{E}\left[ \mathbf{L} \, \mathbf{u} \, (\mathbf{j}^\top \mathbf{L} \, \mathbf{u}) (\mathbf{u}^\top \mathbf{L}^\top \mathbf{j}) \mathbf{j}^\top \right]
                     & = \frac{2}{K^2}\mathbf{L} \, \mathbb{E}\left[ \mathbf{u} \, \mathbf{j}^\top \mathbf{L} \, \mathbf{u} \mathbf{u}^\top \right] \, \mathbf{L}^\top \mathbf{j} \, \mathbf{j}^\top
                     & = \mathbf{0}
    since again the components of the expectation in the next-to-last line involve products of three standard normal random variables.
    Next,
        \mathbf{Q}_3 & = \frac{3}{K^3}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^3 \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top \right]
                     & = \xxx{...}
        \frac{3}{K^3}\mathbf{e} \, \mathbf{j}^\top
    % 3 \, \pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top \, \pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top
        \mathbf{e} \equiv \mathbb{E}\left[
            \mathbf{v}^\top \mathbf{u} \mathbf{u}^\top \mathbf{v}
            \mathbf{v}^\top \mathbf{u}
            \mathbf{L} \mathbf{u}
    and
        \mathbf{v} \equiv \mathbf{L}^\top \, \mathbf{j}
    The components of $\mathbf{e}$ are given by
        e_{i} & =
        \sum\limits_{j,k,l,m}L_{ij} v v_l v_m \mathbb{E}(u_j u u_l u_m)
        \sum\limits_{j,k,l,m}L_{ij} v v_l v_m
        \Big(
        g_4 \, \delta_{j, k} \, \delta_{l, m} \, \delta_{j, l}
        g_2^2 \big(
            \delta_{j, k} \, \delta_{l, m} \, \bar{\delta}_{j, l}
            \delta_{j, l} \, \delta_{k, m} \, \bar{\delta}_{j, k}
            \delta_{j, m} \, \delta_{k, l} \, \bar{\delta}_{j, k}
            \big)
        \Big)
        \xxx{...}
    We may thus write
        \mathbf{e} & = 3 \, \pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top \, \pmb{\Sigma} \, \mathbf{j}
    Inserting this back into the expression for $\mathbf{Q}_3$, we obtain
        \mathbf{Q}_3 & =
        \frac{9}{K^3} \, (\pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top)^2
    Based on this pattern, the general term in $\mathbf{Q}$ is
        \mathbf{Q}_n & =
            \mathbf{0}                                                                                    & n \, \mathrm{even}
            \frac{(n + 1)g_{n+1}}{K^n} \, (\pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top)^\frac{n + 1}{2} & n \, \mathrm{odd}

    \subsection{The $\mathbf{R}$ expectation}
    %
    The first several terms in $\mathbf{R}$ are given by
    %
    \setlength{\abovedisplayskip}{1em}
    \begin{align}
        \mathbf{R}_0 & = \mathbb{E}\left[ \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \mathbf{L}^\top \right]
        \nonumber                                                                                                                                                              \\
                     & = \mathbf{L} \, \mathbb{E}\left[  \mathbf{u} \, \mathbf{u}^\top \right] \, \mathbf{L}^\top
        \nonumber                                                                                                                                                              \\
                     & = \pmb{\Sigma}
        %
        \\[1em]
        %
        \mathbf{R}_1 & = \frac{1}{K}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right) \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \mathbf{L}^\top \right]
        \nonumber                                                                                                                                                              \\
                     & = \mathbf{0}
        %
    \end{align}
    %
    since its elements are all products of three standard normal random variables. Next,
    %
    \begin{align}
        %
        \mathbf{R}_2 & = \mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^2 \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \mathbf{L}^\top \right]
        \nonumber                                                                                                                                                     \\
                     & = \xxx{...}
        \nonumber                                                                                                                                                     \\
                     & = 2 \pmb{\Sigma} \mathbf{j} \, \mathbf{j}^\top \pmb{\Sigma} + \mathrm{sum}(\pmb{\Sigma}) \pmb{\Sigma}
    \end{align}
    %
    \xxx{By trial and error, I found that}
    %
    \begin{align}
        %
        \mathbf{R}_4 & = \frac{5}{K^4}\mathbb{E}\left[ \left(\mathbf{j}^\top \mathbf{L} \, \mathbf{u}\right)^4 \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \mathbf{L}^\top \right]
        \nonumber                                                                                                                                                                  \\
                     & = \xxx{...}
        \nonumber                                                                                                                                                                  \\
                     & = \frac{5}{K^4} \Big( 12 (\pmb{\Sigma} \mathbf{j} \, \mathbf{j}^\top)^2 \pmb{\Sigma} + 3 \, \mathrm{sum}(\pmb{\Sigma})^2 \pmb{\Sigma} \Big)
    \end{align}
    %
    Based on this pattern, the general term in $\mathbf{R}$ is
    %
    \begin{align}
        \mathbf{R}_n & =
        \begin{cases}
            \mathbf{0}                                                                                                                                           & n \, \mathrm{odd}
            \\
            \frac{n (n + 1) g_{n}}{K^n} \, (\pmb{\Sigma} \, \mathbf{j} \, \mathbf{j}^\top)^\frac{n}{2}\pmb{\Sigma} + (n + 1) g_{n} \, m^\frac{n}{2} \pmb{\Sigma} & n \, \mathrm{even}
            \quad.
        \end{cases}
    \end{align}

    \subsection{The final result}
    Inserting the matrices $\mathbf{P}$, $\mathbf{Q}$, and $\mathbf{R}$ into the expression
    for $\tilde{\pmb{\Sigma}}$ and rearranging,
    we obtain
    %
    \begin{align}
        \tilde{\pmb{\Sigma}}
         & =
        \frac{1}{\mu^2} \pmb{\Sigma}
        +
        \frac{1}{\mu^2}
        \sum\limits_{n=1}^\infty
        \frac{(-1)^n(n + 1)}{\mu^{n}}
        \,
        \bigg[
            g_n \, m^\frac{n}{2} \, (\pmb{\Sigma} + \mu^2\mathbf{j} \, \mathbf{j}^\top)
            \, +
            \nonumber                 \\[0.5em]
         & \phantom{XXXXXXXXXXXXXXX.}
        K g_{n+1} \mu  \,
        \left(
        \left(\pmb{\Sigma} \, \frac{\mathbf{j} \, \mathbf{j}^\top}{K^2}\right)^\frac{n + 1}{2}
        +
        \left(\frac{\mathbf{j} \, \mathbf{j}^\top}{K^2} \, \pmb{\Sigma}\right)^\frac{n + 1}{2}
        \right)
        \, +
        \nonumber                     \\[0.5em]
         & \phantom{XXXXXXXXXXXXXXX.}
        n g_n \, \left(\pmb{\Sigma} \, \frac{\mathbf{j} \, \mathbf{j}^\top}{K^2}\right)^\frac{n}{2}\pmb{\Sigma}
        \bigg]
        \quad.
    \end{align}

\else
\fi

>>>>

\section{Non-stationarity}

%
\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/periodic.pdf}
        \oscaption{periodic}{%
            The covariance matrix corresponding to a periodic kernel
            (left) and the covariance matrix of the corresponding
            normalized process (right), plotted on the same color scale.
            In addition to an offset and
            an overall scaling relative to the original, the normalized
            process is noticeably non-stationary.
            \label{fig:periodic}
        }
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/periodic_error.pdf}
        \oscaption{periodic_error}{%
            The covariance of the same normalized process as in
            Figure~\ref{fig:periodic}, but computed numerically
            from $10^5$ samples from the GP
            (left). The right panel shows the relative difference
            between the series solution and the numerical solution,
            normalized to the largest element in $\pmb{\tilde{\Sigma}}$.
            The maximum difference between the two solutions is less
            than one percent and is within the error of the numerical
            estimate.
            \label{fig:periodic_error}
        }
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/periodic_error_asymptotic.pdf}
        \oscaption{periodic_error_asymptotic}{%
            Maximum (blue) and median (orange) normalized relative difference
            between the series solution and the numerical estimate of the
            covariance of the normalized process, as a function of the
            number of samples in the numerical estimate. Thin lines show the
            result of individual trials; thick lines correspond to the average
            over 30 trials.
            \label{fig:periodic_error_asymptotic}
        }
    \end{centering}
\end{figure}

\ifdefined \includeAppendix
    \input{appendix}
\else
\fi
